---
title: "final project"
author: "Yu Huang"
date: "4/26/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## install and load libraries

```{r}
library(lme4)
library(lmerTest)
library(texreg)
library(optimx)
library(lattice)
library(readr)
library(tidyverse)
library(ggplot2)
library(ellipse)
library(nlme)
library(ggpubr)
library(knitr)
library(corrplot)
#install.packages("HLMdiag")
library(HLMdiag)
#install.packages("apaTables")
library(apaTables)
```

## processing data

```{r}
source("all_functions.r")
nyls_data = read_table("r_nyls_data.txt")
# for modeling purposes, remove persons with not enough observations(i.e., < 4 time points)
id_to_del = nyls_data %>%
  count(id) %>%
  filter(n<4) %>%
  select(id)
nyls_data2 = nyls_data %>%
  filter(!id %in% id_to_del$id)
# add integer id called "index"
nyls_data2 = nyls_data2 %>%
  group_by(id) %>%
  mutate(index = cur_group_id())
```

## descriptive statistics
```{r}
summary(nyls_data2)
# summary stats by age
nyls_data2 %>% group_by(age) %>% summarise(M = mean(youth_dev), SD = sd(youth_dev))
apa.1way.table(iv = age, dv = youth_dev, data = nyls_data2, filename = "descriptive stats across age .doc", table.number = 1)
par(mfrow = c(1,3))
hist(x = nyls_data2$youth_dev, col = 2, 
     main = "Histogram of kids' attitude scores",
     xlab = "attitude score")  #left-skewed
hist(x = nyls_data2$age, col = 3,
     main = "Histogram of kids' age",
     xlab = "age")     #right-skewed
hist(x = nyls_data2$parent_att, col = 4,
     main = "Histogram of parents' attitude scores", xlab = "attitude score")      #left-skewed; mostly above 25
par(mfrow = c(1,2))
barplot(table(nyls_data2$gender), col = c("blue", "red"), 
        xlab = "gender", names.arg = c("male", "female"),
        main = "Gender distribution")     #slightly more male kids
barplot(table(nyls_data2$cohort), col = palette.colors(n = 7, "ggplot2"),
        xlab = "cohort", main = "Cohort distribution")    #cohort 7 has smallest number of people
```

### exploratory data analysis
## effects of cohort
# individual data
```{r}
# split data into 7 datasets by cohort
cohort1 = nyls_data2 %>% filter(cohort == 1)
cohort2 = nyls_data2 %>% filter(cohort == 2)
cohort3 = nyls_data2 %>% filter(cohort == 3)
cohort4 = nyls_data2 %>% filter(cohort == 4)
cohort5 = nyls_data2 %>% filter(cohort == 5)
cohort6 = nyls_data2 %>% filter(cohort == 6)
cohort7 = nyls_data2 %>% filter(cohort == 7)
#sample 50 individuals from each cohort for visualizations
cohort1_sample = cohort1 %>% filter(id %in% sample(unique(cohort1$id), size = 50, replace=FALSE))
cohort2_sample = cohort2 %>% filter(id %in% sample(unique(cohort2$id), size = 50, replace = FALSE))
cohort3_sample = cohort3 %>% filter(id %in% sample(unique(cohort3$id), size = 50, replace = FALSE))
cohort4_sample = cohort4 %>% filter(id %in% sample(unique(cohort4$id), size = 50, replace = FALSE))
cohort5_sample = cohort5 %>% filter(id %in% sample(unique(cohort5$id), size = 50, replace = FALSE))
cohort6_sample = cohort6 %>% filter(id %in% sample(unique(cohort6$id), size = 50, replace = FALSE))
cohort7_sample = cohort7 %>% filter(id %in% sample(unique(cohort7$id), size = 50, replace = FALSE))

# Cohort 1
# panel plot with join
par(mfrow = c(1,1))
xyplot(youth_dev ~ age | as.factor(id), data = cohort1_sample,
col.line = "blue", type = c("p","l"),
main = "Plots of Kid's Score by Age (Cohort 1): Join Points")
# panel plot with linear regression
xyplot(youth_dev ~ age | as.factor(id), data = cohort1_sample,
col.line = "red", type = c("p","r"),
main = "Plots of Kids's Score by Age (Cohort 1): Linear Regression ")
# panel plot with spline
xyplot(youth_dev ~ age | as.factor(id), data = cohort1_sample,
col.line = "red", type = c("p","spline"),
main = "Plots of Kids's Score by Age (Cohort 1): Spline")
# score generally decreases with age; nonlinear trend

# Cohort 2
# panel plot with join
xyplot(youth_dev ~ age | as.factor(id), data = cohort2_sample,
col.line = "blue", type = c("p","l"),
main = "Plots of Kid's Score by Age (Cohort 2): Join Points")
# panel plot with linear regression
xyplot(youth_dev ~ age | as.factor(id), data = cohort2_sample,
col.line = "red", type = c("p","r"),
main = "Plots of Kids's Score by Age (Cohort 2): Linear Regression ")
# panel plot with spline
xyplot(youth_dev ~ age | as.factor(id), data = cohort2_sample,
col.line = "red", type = c("p","spline"),
main = "Plots of Kids's Score by Age (Cohort 2): Spline")
# more variability in trend 

# Cohort 3
# panel plot with join
xyplot(youth_dev ~ age | as.factor(id), data = cohort3_sample,
col.line = "blue", type = c("p","l"),
main = "Plots of Kid's Score by Age (Cohort 3): Join Points")
# panel plot with linear regression
xyplot(youth_dev ~ age | as.factor(id), data = cohort3_sample,
col.line = "red", type = c("p","r"),
main = "Plots of Kids's Score by Age (Cohort 3): Linear Regression ")
# panel plot with spline
xyplot(youth_dev ~ age | as.factor(id), data = cohort3_sample,
col.line = "red", type = c("p","spline"),
main = "Plots of Kids's Score by Age (Cohort 3): Spline")
# score generally decreases with age; nonlinear trend 

# Cohort 4
# panel plot with join
xyplot(youth_dev ~ age | as.factor(id), data = cohort4_sample,
col.line = "blue", type = c("p","l"),
main = "Plots of Kid's Score by Age (Cohort 4): Join Points")
# panel plot with linear regression
xyplot(youth_dev ~ age | as.factor(id), data = cohort4_sample,
col.line = "red", type = c("p","r"),
main = "Plots of Kids's Score by Age (Cohort 4): Linear Regression ")
# panel plot with spline
xyplot(youth_dev ~ age | as.factor(id), data = cohort4_sample,
col.line = "red", type = c("p","spline"),
main = "Plots of Kids's Score by Age (Cohort 4): Spline")
# a lot of variability in trend

# Cohort 5
# panel plot with join
xyplot(youth_dev ~ age | as.factor(id), data = cohort5_sample,
col.line = "blue", type = c("p","l"),
main = "Plots of Kid's Score by Age (Cohort 5): Join Points")
# panel plot with linear regression
xyplot(youth_dev ~ age | as.factor(id), data = cohort5_sample,
col.line = "red", type = c("p","r"),
main = "Plots of Kids's Score by Age (Cohort 5): Linear Regression ")
# panel plot with spline
xyplot(youth_dev ~ age | as.factor(id), data = cohort5_sample,
col.line = "red", type = c("p","spline"),
main = "Plots of Kids's Score by Age (Cohort 5): Spline")
# a lot of variability in trend

# Cohort 6
# panel plot with join
xyplot(youth_dev ~ age | as.factor(id), data = cohort6_sample,
col.line = "blue", type = c("p","l"),
main = "Plots of Kid's Score by Age (Cohort 6): Join Points")
# panel plot with linear regression
xyplot(youth_dev ~ age | as.factor(id), data = cohort6_sample,
col.line = "red", type = c("p","r"),
main = "Plots of Kids's Score by Age (Cohort 6): Linear Regression ")
# panel plot with spline
xyplot(youth_dev ~ age | as.factor(id), data = cohort6_sample,
col.line = "red", type = c("p","spline"),
main = "Plots of Kids's Score by Age (Cohort 6): Spline")
# a lot of variability in trend

# Cohort 7
# panel plot with join
xyplot(youth_dev ~ age | as.factor(id), data = cohort7_sample,
col.line = "blue", type = c("p","l"),
main = "Plots of Kid's Score by Age (Cohort 7): Join Points")
# panel plot with linear regression
xyplot(youth_dev ~ age | as.factor(id), data = cohort7_sample,
col.line = "red", type = c("p","r"),
main = "Plots of Kids's Score by Age (Cohort 7): Linear Regression ")
# panel plot with spline
xyplot(youth_dev ~ age | as.factor(id), data = cohort7_sample,
col.line = "red", type = c("p","spline"),
main = "Plots of Kids's Score by Age (Cohort 7): Spline")
# a lot of variability in trend

# Join points by subject
par(mfrow=c(1,1))
join_by_sub = function(df, name){
  df$id = as.factor(df$id)
  plot(df$age, df$youth_dev, type = 'n', 
       cex.main = 1.5,
       xlab = 'Age',
       ylab = "Kid's Score",
       main = paste("Join points for Person: ", name)
  )
  ids = unique(df$id)
  for (j in seq_along(ids)) {
    sub = df %>% filter(id == ids[j])
    lines(sub$age,sub$youth_dev,col=j)
  }
}
join_by_sub(cohort1_sample, name = "Cohort 1 Sample") # variable, nonlinear trend w/ increasing variance
join_by_sub(cohort2_sample, name = "Cohort 2 Sample") # variable, nonlinear trend w/ increasing variance
join_by_sub(cohort3_sample, name = "Cohort 3 Sample") # variable, nonlinear trend w/ increasing variance
join_by_sub(cohort4_sample, name = "Cohort 4 Sample") # variable, nonlinear trend
join_by_sub(cohort5_sample, name = "Cohort 5 Sample") # variable, nonlinear trend
join_by_sub(cohort6_sample, name = "Cohort 6 Sample") # variable, nonlinear trend
join_by_sub(cohort7_sample, name = "Cohort 7 Sample") # variable, nonlinear trend

# linear regression by subject
linear_by_sub = function(df, name){
  df$id = as.factor(df$id)
  plot(df$age, df$youth_dev, type = 'n', 
       cex.main = 1.5,
       xlab = 'Age',
       ylab = "Kid's Score",
       main = paste("Separate Linear Regression for Person: ", name)
  )
  ids = unique(df$id)
  for (j in seq_along(ids)) {
    sub = df %>% filter(id == ids[j])
    fitted = fitted(lm(youth_dev~age,sub))
    lines(sub$age,fitted,col=j)
  }
}
linear_by_sub(cohort1_sample, "Cohort 1 Sample") # more negative slopes
linear_by_sub(cohort2_sample, "Cohort 2 Sample") # variable slopes
linear_by_sub(cohort3_sample, "Cohort 3 Sample") # more negative slopes
linear_by_sub(cohort4_sample, "Cohort 4 Sample") # variable slopes
linear_by_sub(cohort5_sample, "Cohort 5 Sample") # variable slopes
linear_by_sub(cohort6_sample, "Cohort 6 Sample") # variable slopes
linear_by_sub(cohort7_sample, "Cohort 7 Sample") # variable slopes

# linear regression w/ quadratic by subject
quad_by_sub = function(df, name){
  df$id = as.factor(df$id)
  plot(df$age, df$youth_dev, type = 'n', 
       cex.main = 1.5,
       xlab = 'Age',
       ylab = "Kid's Score",
       main = paste("Separate Quadratic Regression for Person: ", name)
  )
  ids = unique(df$id)
  for (j in seq_along(ids)) {
    sub = df %>% filter(id == ids[j])
    sub$agesq = sub$age * sub$age
    fitted = fitted(lm(youth_dev~age + agesq,sub))
    lines(sub$age,fitted,col=j)
  }
}
quad_by_sub(cohort1_sample, "Cohort 1 Sample")
quad_by_sub(cohort2_sample, "Cohort 2 Sample")
quad_by_sub(cohort3_sample, "Cohort 3 Sample")
quad_by_sub(cohort4_sample, "Cohort 4 Sample")
quad_by_sub(cohort5_sample, "Cohort 5 Sample")
quad_by_sub(cohort6_sample, "Cohort 6 Sample")
quad_by_sub(cohort7_sample, "Cohort 7 Sample")
```

# mean data
```{r}
# Mean by age
mean_youth_dev_by_age = nyls_data2 %>%
  group_by(age) %>%
  summarise(mean_youth_dev = mean(youth_dev))
par(mfrow = c(1,1))
plot(mean_youth_dev_by_age$age, mean_youth_dev_by_age$mean_youth_dev,
     type = 'b', lwd = 2, col = "blue",
     main = "Overall Mean Kid's Score by Age",
     xlab = "Age", ylab = "Score")
# score decreases first and slightly increases at 20. steepest decrease from 13 to 14. nonlinear relationship btw score and age.
# Mean by age by cohort
mean_youth_dev_by_age_cohort = nyls_data2 %>%
  group_by(age, cohort) %>%
  summarise(mean_youth_dev = mean(youth_dev)) %>%
  arrange(cohort)
ggplot(data = mean_youth_dev_by_age_cohort, aes(x = age, y = mean_youth_dev, group = factor(cohort))) + geom_line(aes(colour = factor(cohort))) + geom_point(aes(color = factor(cohort))) + labs(title = "Mean Kid's Score by Age \n Separate Lines for cohorts", x = "Age", y = "Score")
# score decreases with age more steeply for younger cohorts (cohort 1,2,3) and cohort 5. from cohort 5 on, score decreases slower and eventually levels off around 18-19. older cohorts (cohort 6,7) tend to have similar starting scores, but their ending scores are reversely ordered (cohort 6 < cohort 7).
```

## effects of gender
```{r}
males = nyls_data2 %>% filter(gender == 1)
females = nyls_data2 %>% filter(gender == 2)
sampled_ids_male = sample(unique(males$id), size = 50, replace = FALSE)
males_sample = males %>% filter(id %in% sampled_ids_male)
sampled_ids_female = sample(unique(females$id), size = 50, replace = FALSE)
females_sample = females %>% filter(id %in% sampled_ids_female)
join_by_sub(males_sample, "Male Sample")    # nonlinear trend w/ increasing variance
linear_by_sub(males_sample, "Male Sample")  # more negative slopes
join_by_sub(females_sample, "Female Sample")   # nonlinear trend w/ increasing variance
linear_by_sub(females_sample, "Female Sample")  # mix of negative and positive slopes; more positive slopes than males
# Mean by age by gender
mean_youth_dev_by_age_gender = nyls_data2 %>%
  group_by(age, gender) %>%
  summarise(mean_youth_dev = mean(youth_dev))
ggplot(data = mean_youth_dev_by_age_gender, aes(x = age, y = mean_youth_dev, group = factor(gender))) + geom_line(aes(colour = factor(gender))) + geom_point(aes(color = factor(gender))) + labs(title = "Mean Kid's Score by Age \n Separate Lines for Males and Females", x = "Age", y = "Score") + scale_color_manual(values = c("1" = "blue", "2" = "red"), labels = c("Male", "Female"))
# males on average score lower than females at any age, and the difference in avg scores becomes larger as age increases. male scores decrease faster than female scores overall. scores start to increase for both genders after 18, earlier for females.
```

## effects of parents' attitude
```{r}
# group kids based on their parents' attitude score;for each group, plot kid's score by age
xyplot(youth_dev ~ age | as.factor(parent_att), data = nyls_data2,
col.line = "blue", type = c("p","spline"),
main = "Plots of Kid's Score by Age: Parents' Attitude")
# for parents' attitude score of 27-29 and 31-35, similar trend in kid's score by age; for parents' attitude score of 30 and 31, more linear trend in kid's score by age; low parents' attitude scores have variable trends, but this may be due to small sample sizes.
```

## Rsq's and metaRsq
```{r}
# model 1: linear
ssmodel = c(0)
sstotal = c(0)
id = unique(nyls_data2$id)
R2 = matrix(99, nrow = length(id), ncol = 2)
for (i in 1:length(id)){
  sub = nyls_data2 %>% filter(index == i)
  model0 = lm(youth_dev ~ age, data = sub)
  a = anova(model0)
  ssmodel = ssmodel + a[1,2]
  sstotal = sstotal + sum(a[,2])
  R2[i,1:2] = c(i, summary(model0)$r.squared)
}
R2meta.mod1 = ssmodel/sstotal   # meta R^2 for model 1
R2.mod1 = R2[,2]   # R^2 for each individual 
R2.mod1 = as.data.frame(R2.mod1)
ggplot(R2.mod1, aes(x = R2.mod1)) + geom_histogram(aes(y=..density..), colour = "black", fill = "white") + geom_density(alpha = 0.2, fill = "#FF6666") + geom_vline(aes(xintercept=R2meta.mod1), colour = "red", size = 1) + geom_text(aes(x = R2meta.mod1, label = "R^2_meta", y=2), colour = "black") + labs(title = expression(paste(R^2," Model 1:   ",Y[it], " = ", beta[oi], " + ", beta[li], x[lit], " + ", R[it])), x = expression(R^2))

# model 2: quadratic
ssmodel = c(0)
sstotal = c(0)
R2 = matrix(99, nrow = length(id), ncol = 2)
for (i in 1:length(id)){
  sub = nyls_data2 %>% filter(index == i)
  sub$agesq = sub$age * sub$age
  model1 = lm(youth_dev ~ age + agesq, data = sub)
  a = anova(model1)
  ssmodel = ssmodel + a[1,2] + a[2,2]
  sstotal = sstotal + sum(a[,2])
  R2[i,1:2] = c(i, summary(model1)$r.squared)
}
R2meta.mod2 = ssmodel/sstotal   # meta R^2 for model 2 (improved a lot)
R2.mod2 = R2[,2]   # R^2 for each individual 
R2.mod2 = as.data.frame(R2.mod2)
ggplot(R2.mod2, aes(x = R2.mod2)) + geom_histogram(aes(y=..density..), colour = "black", fill = "white") + geom_density(alpha = 0.2, fill = "#FF6666") + geom_vline(aes(xintercept=R2meta.mod2), colour = "red", size = 1) + geom_text(aes(x = R2meta.mod2, label = "R^2_meta", y=2), colour = "black") + labs(title = expression(paste(R^2," Model 2:   ",Y[it], " = ", beta[oi], " + ", beta[li], x[lit], " + ", beta[q,i], x[qit]^2, " + " ,R[it])), x = expression(R^2))

### comparison: who is improving
Rsqs = bind_cols(R2.mod1, R2.mod2)
Rsqs$index = 1:length(id)
combined = nyls_data2 %>% left_join(Rsqs, by = "index")
ggplot(combined, aes(x = R2.mod1, y = R2.mod2, group = factor(cohort))) +
  geom_point(aes(colour = factor(cohort))) + geom_abline(slope = 1, intercept = 0, col = "black") + labs(title = "Who is better fit by quadratic term ", 
                                                                                                       y = expression(paste(R^2," Model 2:   ",Y[it], " = ", beta[oi], " + ", beta[li], x[lit], " + ", beta[q,i], x[qit]^2, " + " ,R[it])),
                                                                                                       x = expression(paste(R^2," Model 1:   ",Y[it], " = ", beta[oi], " + ", beta[li], x[lit], " + " ,R[it])))
# no discernible performance differences among cohorts
ggplot(combined, aes(x = R2.mod1, y = R2.mod2, group = factor(gender))) +
  geom_point(aes(colour = factor(gender))) + geom_abline(slope = 1, intercept = 0, col = "black") + labs(title = "Who is better fit by quadratic term ", 
                                                                                                       y = expression(paste(R^2," Model 2:   ",Y[it], " = ", beta[oi], " + ", beta[li], x[lit], " + ", beta[q,i], x[qit]^2, " + " ,R[it])),
                                                                                                       x = expression(paste(R^2," Model 1:   ",Y[it], " = ", beta[oi], " + ", beta[li], x[lit], " + " ,R[it]))) + scale_color_manual(values = c("1" = "blue", "2" = "red"), labels = c("Male","Female"))
# no discernible performance differences between genders
```

## exploring random effects
# Compute mean squared residual from lm and fit regression to them
# Looking at variance functions by looking at
#  (mean) squared residuals
```{r}
nyls_data2$gender = as.factor(nyls_data2$gender)
nyls_data2$cohort = as.factor(nyls_data2$cohort)
nyls_data2$agesq = nyls_data2$age**2
ols = lm(youth_dev ~ 1 + age + agesq + gender + cohort + parent_att +
           age*gender + age*cohort, data = nyls_data2)
summary(ols)
nyls_data2$res.sq = residuals(ols)*residuals(ols)
mres_cohort = nyls_data2 %>%
  group_by(age, cohort) %>%
  summarise(mean_res.sq = mean(res.sq))
p1 = ggplot(mres_cohort, aes(x=age, y=mean_res.sq, group = cohort)) + geom_line(aes(colour = cohort)) + labs(title = "Data", x = "Age", y = "Mean Squared Residuals")

lin_fitted = matrix(0, nrow = 7, ncol = 4)
for (i in 1:7){
  mod = lm(mean_res.sq ~ age, data = mres_cohort %>% filter(cohort == i))
  lin_fitted[i,] = fitted(mod) 
}
lin_fitted = as.data.frame(lin_fitted)
colnames(lin_fitted) = c("age1","age2","age3","age4")
lin_fitted = lin_fitted %>%
  pivot_longer(cols = age1:age4, names_to = "age", values_to = "fitted.lin")
mres_cohort = mres_cohort %>% arrange(cohort)
mres_cohort$fitted_lin = lin_fitted$fitted.lin
p2 = ggplot(mres_cohort, aes(x=age, y=fitted_lin, group = cohort)) + geom_line(aes(colour = cohort)) + labs(title = "Fit using Linear Regression", x = "Age", y = "Mean Squared Residuals")

quad_fitted = matrix(0, nrow = 7, ncol = 4)
for (i in 1:7){
  mres_cohort$age2 = mres_cohort$age*mres_cohort$age
  mod = lm(mean_res.sq ~ age + age2, data = mres_cohort %>% filter(cohort == i))
  quad_fitted[i,] = fitted(mod)
}
quad_fitted = as.data.frame(quad_fitted)
quad_fitted = quad_fitted %>%
  pivot_longer(cols = V1:V4, names_to = "age", values_to = "fitted.quad")
mres_cohort$fitted_quad = quad_fitted$fitted.quad
p3 = ggplot(mres_cohort, aes(x=age, y=fitted_quad, group = cohort)) + geom_line(aes(colour = cohort)) + labs(title = "Quadratic", x = "Age", y = "Mean Squared Residuals")

cubic_fitted = matrix(0, nrow = 7, ncol = 4)
for (i in 1:7){
  mres_cohort$age3 = mres_cohort$age2*mres_cohort$age
  mod = lm(mean_res.sq ~ age + age2 + age3, data = mres_cohort %>% filter(cohort == i))
  cubic_fitted[i,] = fitted(mod)
}
cubic_fitted = as.data.frame(cubic_fitted)
cubic_fitted = cubic_fitted %>%
  pivot_longer(cols = V1:V4, names_to = "age", values_to = "fitted.cubic")
mres_cohort$fitted_cubic = cubic_fitted$fitted.cubic
p4 = ggplot(mres_cohort, aes(x=age, y=fitted_cubic, group = cohort)) + geom_line(aes(colour = cohort)) + labs(title = "Cubic", x = "Age", y = "Mean Squared Residuals")

ggarrange(p1,p2,p3,p4, ncol = 2, nrow = 2)  # cubic fit is a perfect fit (since there are only 4 observations per cohort)
```

# looking at serial correlation
```{r}
create_cor_tab = function(i){
  data = nyls_data2 %>% filter(cohort == i) %>% select(id, age, youth_dev) %>% pivot_wider(names_from = age, names_prefix = "age_", values_from = youth_dev)
  ctab = cor(data[,-1], use = "pairwise.complete.obs")
  ctab
}
nyls_data3 = nyls_data2 %>% select(id, age, youth_dev) %>% arrange(age) %>%
  pivot_wider(names_from = age, names_prefix = "age_", values_from = youth_dev)
ctab_all = cor(nyls_data3[,-1], use = "pairwise.complete.obs")
corrplot(ctab_all, method = "number")

cohort1_ctab = create_cor_tab(i = 1)
#kable(cohort1_ctab, caption = "Residual Correlation Matrix for Cohort 1 ")
cohort2_ctab = create_cor_tab(i = 2)
cohort3_ctab = create_cor_tab(i = 3)
cohort4_ctab = create_cor_tab(i = 4)
cohort5_ctab = create_cor_tab(i = 5)
cohort6_ctab = create_cor_tab(i = 6)
cohort7_ctab = create_cor_tab(i = 7)

par(mfrow = c(2,4))
corrplot(cohort1_ctab, method = "number")
corrplot(cohort2_ctab, method = "number")
corrplot(cohort3_ctab, method = "number")
corrplot(cohort4_ctab, method = "number")
corrplot(cohort5_ctab, method = "number")
corrplot(cohort6_ctab, method = "number")
corrplot(cohort7_ctab, method = "number")
```

### Modeling
## No serial correlation
```{r}
# Null/empty Model w/ random intercept 
mod0 = lmer(youth_dev ~ 1 + (1 | id), data = nyls_data2, REML=FALSE, control = lmerControl(optimizer ="Nelder_Mead"))
summary(mod0)
icc(mod0)

# main effects of age, age^2, gender, cohort, and parent_att w/ random intercept
mod1 = lmer(youth_dev ~ 1 + age + agesq + gender + cohort + parent_att + (1 | id), data = nyls_data2, REML=FALSE, control = lmerControl(optimizer = "Nelder_Mead"))
summary(mod1)
icc(mod1)

# add cross-level interaction effects for age*gender and age*cohort
mod2 = lmer(youth_dev ~ 1 + age + agesq + gender + cohort + parent_att +
              age*gender + age*cohort + (1 | id), data = nyls_data2, REML=FALSE, control = lmerControl(optimizer = "Nelder_Mead"))
summary(mod2)
icc(mod2)

# add random slope for age
mod3 = lmer(youth_dev ~ 1 + age + agesq + gender + cohort + parent_att +
              age*gender + age*cohort + (1 + age | id), data = nyls_data2, REML=FALSE, control = lmerControl(optimizer = "Nelder_Mead"))

# check if lack of convergence is OK to ignore
relgrad = with(mod3@optinfo$derivs,solve(Hessian,gradient))
max(abs(relgrad)) #<0.001. OK
summary(mod3)

# add random slope for age^2
mod4 = lmer(youth_dev ~ 1 + age + agesq + gender + cohort + parent_att +
              age*gender + age*cohort + (1 + age + agesq| id), data = nyls_data2, REML=FALSE, control = lmerControl(optimizer = "Nelder_Mead"))
# rescale age^2
nyls_data2$xagesq = nyls_data2$agesq/sd(nyls_data2$agesq)
sd(nyls_data2$xagesq)
mod4_rescaled = lmer(youth_dev ~ 1 + age + xagesq + gender + cohort + parent_att +
              age*gender + age*cohort + (1 + age + xagesq| id), data = nyls_data2, REML=FALSE, control = lmerControl(optimizer = "Nelder_Mead"))
# check if lack of convergence is OK
relgrad = with(mod4_rescaled@optinfo$derivs,solve(Hessian,gradient))
max(abs(relgrad)) # <0.001.OK
summary(mod4_rescaled)

htmlreg(list(mod0, mod1, mod2, mod3, mod4_rescaled),
        custom.model.names = c("Null", "All Main Effects", "+ Some Cross Effects", "+ Random Effect for Age", "+ Random Effect for Age^2"), single.row = TRUE, file = "model0-model4.doc")  # model 4 is the preliminary HLM


# model reduction: random effects
# test whether need random slope for age^2 in preliminary model
lr = deviance(mod3) - deviance(mod4_rescaled)
df0 = 2  #random intercept and 1 random slope
df1 = 3  #random intercept and 2 random slopes
p1 = pchisq(lr, df1, lower.tail = FALSE)
p0 = pchisq(lr, df0, lower.tail = FALSE)
pvalue = 0.5*(p1+p0)
pvalue    # reject null. need random slope for age^2

# model reduction: fixed effects
# robust estimation for standard errors of fixed effects
robust_table = 
round(robust(mod4_rescaled, nyls_data2$youth_dev, nyls_data2$id, "between/within"),digits = 4)   #same conclusion as model-based estimation for each fixed effect
write.csv(robust_table, "robust_se.csv")
robust_table2 =
round(robust(mod4_rescaled, nyls_data2$youth_dev, nyls_data2$id, "satterthwaite"),digits = 4)  
write.csv(robust_table2, "robust_se_2.csv")
#same conclusion as model-based estimation for each fixed effect. t-tests indicate don't need fixed effects for cohorts or cross-level interaction between age and cohort; maybe don't need fixed effect of gender.
# contrast test for whether need main effects and cross-level interaction effects involving cohort: 
# test cohort2 = cohort3 = cohort4 = cohort5 = cohort6 = cohort7 = 0 and all interactions involving cohort = 0 
L = matrix(0, nrow=12, ncol=18)
for (i in 1:6){
  L[i,i+4] = 1
  L[i+6,i+12] =1
}
contrast(mod4_rescaled, L)  # retain null. remove effects for cohort.

mod5 = lmer(youth_dev ~ 1 + age + xagesq + gender + parent_att +
              age*gender + (1 + age + xagesq| id), data = nyls_data2, REML=FALSE,
            control = lmerControl(optimizer = "nlminbwrap"))
relgrad = with(mod5@optinfo$derivs,solve(Hessian,gradient))
max(abs(relgrad)) #<0.001. OK.
summary(mod5)
# likelihood ratio test for effects for cohort
anova(mod5, mod4_rescaled)  # support for removing effects for cohort.
# compare global fit measures of mod5 with mod4_rescaled
bic.hlm(mod4_rescaled, cluster.id = nyls_data2$id)
bic.hlm(mod5, cluster.id = nyls_data2$id)   # the reduced model is better in terms of every measure. use mod5 as the final model
bic.hlm(mod0, cluster.id = nyls_data2$id)

htmlreg(list(mod4_rescaled, mod5),
        custom.model.names = c("Preliminary", "No Cohort"), single.row = TRUE,
        file = "mod4-mod5.doc")
htmlreg(list(mod5), custom.model.names = "No Cohort", single.row = TRUE, file = "mod5.doc")
```

## With Serial Correlation
```{r}
# built on the final model we chose above
# empty/null model w/o serial correlation: mod0
nlme0 = lme(youth_dev ~ 1, random = ~1 |id, data = nyls_data2)
summary(nlme0)
# more fixed effects w/o serial correlation
nlme1 = lme(youth_dev ~ 1 + age + xagesq + gender + parent_att + gender*age, 
              random = ~ 1|id, data = nyls_data2)
summary(nlme1)
# AR(1) w/ random intercept and slope
nlme2 = lme(youth_dev ~ 1 + age + xagesq + gender + parent_att + gender*age,
            random = ~ 1 + age + xagesq | id,
            correlation = corAR1(form = ~ 1 + age | id),
            data = nyls_data2, control = lmeControl(msMaxIter = 1000, msMaxEval = 1000))
# error: system is computationally singular. drop random slope for agesq
nlme2 = lme(youth_dev ~ 1 + age + xagesq + gender + parent_att + gender*age,
            random = ~ 1 + age | id,
            correlation = corAR1(form = ~ 1 + age | id),
            data = nyls_data2, control = lmeControl(msMaxIter = 1000, msMaxEval = 1000))
# nlminb problem, convergence error code = 1
# message = singular convergence (7)
# drop random slope for age (AR(1) w/ random intercept only)
nlme2 = lme(youth_dev ~ 1 + age + xagesq + gender + parent_att + gender*age,
            random = ~ 1 | id,
            correlation = corAR1(form = ~ 1 + age | id),
            data = nyls_data2, control = lmeControl(msMaxIter = 1000, msMaxEval = 1000))
summary(nlme2)
screenreg(list(mod5, nlme2))    # AIC, BIC worse than mod5

# MA(1) w/ random intercept only
nlme3 = lme(youth_dev ~ 1 + age + xagesq + gender + parent_att + gender*age,
            random = ~ 1 | id,
            correlation = corARMA(form = ~ 1 + age | id, p = 0, q = 1),
            data = nyls_data2, control = lmeControl(msMaxIter = 1000, msMaxEval = 1000))
summary(nlme3)
screenreg(list(mod5,nlme2, nlme3))  #AIC, BIC worse than AR(1) w/ random intercept

# ARMA(1,1) w/ random intercept only
nlme4 = lme(youth_dev ~ 1 + age + xagesq + gender + parent_att + gender*age,
            random = ~ 1 | id,
            correlation = corARMA(form = ~ 1 + age | id, p = 1, q = 1),
            data = nyls_data2, control = lmeControl(msMaxIter = 1000, msMaxEval = 1000))
summary(nlme4)
screenreg(list(mod5,nlme2, nlme3, nlme4)) #best AIC,BIC among models w/ serial correlation; but still worse than mod5

# drop fixed and random effects for agesq and add AR(1) w/ random intercept and slope for age
nlme5 = lme(youth_dev ~ 1 + age + gender + parent_att + gender*age,
            random = ~ 1 + age | id,
            correlation = corAR1(form = ~ 1 + age | id),
            data = nyls_data2, control = lmeControl(msMaxIter = 1000, msMaxEval = 1000))
summary(nlme5)
screenreg(list(mod5,nlme2, nlme3, nlme4,nlme5))

# drop random intercept
nlme6 = lme(youth_dev ~ 1 + age + gender + parent_att + gender*age,
            random = ~ -1 + age | id,
            correlation = corAR1(form = ~ 1 + age | id),
            data = nyls_data2, control = lmeControl(msMaxIter = 1000, msMaxEval = 1000))
anova(nlme6, nlme5)
lr = 6.101145
p1 = 1-pchisq(lr, df = 2)
p2 = 1-pchisq(lr, df = 1)
pvalue = 0.5*(p1+p2)
pvalue     # reject null. keep random intercept. 
screenreg(list(nlme2, nlme3, nlme4,nlme5))
htmlreg(list(mod0, mod5, nlme2, nlme3, nlme4, nlme5, nlme6), 
        custom.model.names = c("Null", "Model 5", "AR1 w/ U_0i", "MA1 w/ U_0i", "ARMA(1,1) w/ U_0i", "AR1 w/ U_0i and U1_i", "AR1 w/ U_1i"),
        single.row = TRUE, file = "serial_corr mods.doc"
        )

intervals(nlme6, method = profile) # altho fixed effect of gender appears to be non-significant, it is significant for cross effect with age; so don't drop it

###############################################################
# Final Model: random intercept only, ARMA(1)                 #
###############################################################
# get model-based COV and CORR matrices for Y
tau0sq = 2.239857**2
z = matrix(1, nrow = 4, ncol = 4)
between = tau0sq*z

rho = 0.7084814
theta = -0.3294492
sigmasq = nlme4$sigma**2
omega = diag(4)*(1+theta**2-2*rho*theta)
for(i in 1:3){
  for(j in (i+1):4){
    power = j-i-1
    omega[i,j] = (1-rho*theta)*(rho-theta)*(rho**power)
    omega[j,i] = omega[i,j]
  }
}
within = sigmasq/(1-rho**2)*omega
covY = between + within
write.csv(covY, "COV_ARMA(1,1).csv")

isd = solve(diag(sqrt(diag(covY)))) 
Ry = isd %*% covY %*% isd
write.csv(Ry, "CORR_ARMA(1,1).csv")
```

Based on global fit measures, I chose to use model 5 as my final model.

## model diagnostics
# level 1 residual plots
```{r}
res_level1 = resid(mod5)
fit = predict(mod5)   # conditional fitted values
par(mfrow = c(2,2))
plot(fit, res_level1, 
     xlab='Conditional Fitted Values',
     ylab='Standardized Pearson Residuals',
     main='Conditional Residuals')

qqnorm(res_level1)
qqline(res_level1, col = "blue")

h = hist(res_level1, breaks = 20)
xfit = seq(min(res_level1), max(res_level1), length=50)
yfit = dnorm(xfit, mean=mean(res_level1), sd=sd(res_level1))
yfit = yfit*diff(h$mids[1:2])*length(res_level1) # use mid-points 
lines(xfit, yfit, col='blue', lwd=2)

plot.new()
text(0.5, 0.8, "Model 5")
text(0.5,0.5, 'Deviance=30186')
text(0.5,0.2, 'AIC=30212')
```

# qq plots for U0j, U1j, and U2j
```{r}
ranU = as.data.frame(ranef(mod5))
U0j = ranU[which(ranU$term=="(Intercept)"),]
U1j = ranU[which(ranU$term=="age"),]
U2j = ranU[which(ranU$term=="xagesq"),]
U0j = U0j[order(U0j$condval),]
U1j = U1j[order(U1j$condval),]
U2j = U2j[order(U2j$condval),]
U0j$grp = seq(1:1424)
U1j$grp = seq(1:1424)
U2j$grp = seq(1:1424)
# qq plot for U0j
par(mfrow=c(1,1))
qqdata = qqnorm(U0j$condval)
plot(qqdata$x,qqdata$y,type="p",pch=19,col="red",
main="Normal QQ plot of U0j with 95% Confidence Bands",
     xlab="Theoretical Value",
     ylab="Estimated U0j")
arrows(qqdata$x, qqdata$y - 1.96*U0j$condsd, 
       qqdata$x, qqdata$y + 1.96*U0j$condsd, 
       length = 0.00, col = rgb(0, 0, 1, 0.2))
qqline(U0j$condval,col="yellow")
# for U1j
qqdata = qqnorm(U1j$condval)
plot(qqdata$x,qqdata$y,type="p",pch=19,col="red",
main="Normal QQ plot of Uuj with 95% Confidence Bands",
     xlab="Theoretical Value",
      ylab="Estimated U1j")
arrows(qqdata$x, qqdata$y - 1.96*U1j$condsd, 
       qqdata$x, qqdata$y + 1.96*U1j$condsd, 
	length=0.00, col = rgb(0, 0, 1, 0.2))
qqline(U1j$condval,col="yellow")
# for U2j
qqdata = qqnorm(U2j$condval)
plot(qqdata$x,qqdata$y,type="p",pch=19,col="red",
main="Normal QQ plot of Uuj with 95% Confidence Bands",
     xlab="Theoretical Value",
      ylab="Estimated U2j")
arrows(qqdata$x, qqdata$y - 1.96*U2j$condsd, 
       qqdata$x, qqdata$y + 1.96*U2j$condsd, 
	length=0.00, col = rgb(0, 0, 1, 0.2))
qqline(U2j$condval,col="yellow")
```

# level 2 residual plots
```{r}
res_final = hlm_resid(mod5, level = 'id', type = 'EB', standardize = TRUE,
          include.ls = FALSE)
head(res_final)

par(mfrow = c(1,3))
# histograms of level 2 residuals
# U0_i
h1 = hist(res_final$.std.ranef.intercept, breaks = 20, main = "Histogram of U0_j", xlab = "U0_j")
xfit <- seq(min(res_final$.std.ranef.intercept), max(res_final$.std.ranef.intercept), length=50)
yfit <- dnorm(xfit, mean=mean(res_final$.std.ranef.intercept), sd=sd(res_final$.std.ranef.intercept))
yfit <- yfit*diff(h1$mids[1:2])*length(res_final$id) # use mid-points 
lines(xfit, yfit, col='blue', lwd=2)             # draws normal

# U1_i
h2 = hist(res_final$.std.ranef.age, breaks = 20, main = "Histogram of U1_j", xlab = "U1_j")
xfit <- seq(min(res_final$.std.ranef.age), max(res_final$.std.ranef.age), length=50)
yfit <- dnorm(xfit, mean=mean(res_final$.std.ranef.age), sd=sd(res_final$.std.ranef.age))
yfit <- yfit*diff(h2$mids[1:2])*length(res_final$id) # use mid-points 
lines(xfit, yfit, col='blue', lwd=2)             # draws normal

# U2_i
h3 = hist(res_final$.std.ranef.xagesq, breaks = 20, main = "Histogram of U2_j", xlab = "U2_j")
xfit <- seq(min(res_final$.std.ranef.xagesq), max(res_final$.std.ranef.xagesq), length=50)
yfit <- dnorm(xfit, mean=mean(res_final$.std.ranef.xagesq), sd=sd(res_final$.std.ranef.xagesq))
yfit <- yfit*diff(h3$mids[1:2])*length(res_final$id) # use mid-points 
lines(xfit, yfit, col='blue', lwd=2)             # draws normal
```

